---
title: "International Math Olympiad (IMO) Data Visualizations"
format: html
execute: 
  echo: false
  warning: false
  message: false
embed-resources: true
---


```{r}
#| label: setup
#| include: false
library(tidyverse)
library(purrr)
```
# Data Description

Data used in this project was sourced from the [Tidy Tuesday](https://github.com/rfordatascience/tidytuesday) Github repository. The specific data and follwing description used can be found here: 
[2024-09-24: IMO Data](https://github.com/rfordatascience/tidytuesday/tree/main/data/2024/2024-09-24). 

> The International Mathematical Olympiad (IMO) is the World Championship Mathematics Competition for High School students and is held annually in a different country. The first IMO was held in 1959 in Romania, with 7 countries participating. It has gradually expanded to over 100 countries from 5 continents. The competition consists of 6 problems and is held over two consecutive days with 3 problems each.


# Data Cleaning

> 

```{r}
IMO_data <- read_csv("project-files/data-raw/individual_results_df.csv")
country_continents <- read_csv("project-files/data-raw/list-of-countries-by-continent-2025.csv")
```

```{r}
fig_1_clean_data <- IMO_data %>% 
  select(p1, p2, p3, p4, p5, p6, p7, year) %>% 
  group_by(year) %>% 
  summarise(p1 = mean(p1, na.rm = TRUE),
            p2 = mean(p2, na.rm = TRUE),
            p3 = mean(p3, na.rm = TRUE),
            p4 = mean(p4, na.rm = TRUE),
            p5 = mean(p5, na.rm = TRUE),
            p6 = mean(p6, na.rm = TRUE)) %>% 
  pivot_longer(cols = 2:7, names_to = "problem_num", values_to = "mean_score")
```

```{r}
write_csv(fig_1_clean_data, "project-files/data-clean/fig_1_clean_data.csv")
```

```{r}
fig_2_clean_data <- IMO_data %>% 
  left_join(country_continents, by = "country") %>% 
  mutate(continent = case_when(country == "People's Republic of China" ~ "Asia",
                               country == "United States of America" ~ "North America",
                               country == "Republic of Korea" ~ "Asia",
                               country == "Türkiye" ~ "Asia",
                               country == "Islamic Republic of Iran" ~ "Asia",
                               country == "Republic of Moldova" ~ "Europe",
                               country == "Kosovo" ~ "Europe",
                               country == "Russian Federation" ~ "Asia",
                               country == "Democratic People's Republic of Korea" ~ "Asia",
                               country == "Serbia and Montenegro" ~ "Europe",
                               country == "Yugoslavia" ~ "Europe",
                               country == "Turkish Republic of Northern Cyprus" ~ "Asia",
                               country == "Yugoslavia" ~ "Europe",
                               country == "Czechoslovakia" ~ "Europe",
                               country == "Union of Soviet Socialist Republics" ~ "Asia",
                               country == "German Democratic Republic" ~ "Asia",
                               continent == "Oceania" ~ "Australia",
                              TRUE ~ continent)) %>% 
  filter(!is.na(continent)) %>% 
  select(p1, p2, p3, p4, p5, p6, p7, year, continent) %>% 
  group_by(continent, year) %>% 
  summarise(p1 = mean(p1, na.rm = TRUE),
            p2 = mean(p2, na.rm = TRUE),
            p3 = mean(p3, na.rm = TRUE),
            p4 = mean(p4, na.rm = TRUE),
            p5 = mean(p5, na.rm = TRUE),
            p6 = mean(p6, na.rm = TRUE)) %>% 
  pivot_longer(cols = 3:8, names_to = "problem_num", values_to = "mean_score") %>% 
  mutate(mean_score = case_when(mean_score == "NaN" ~ NA,
                                TRUE ~ mean_score)) %>% 
  filter(continent != "Oceania")
```

```{r}
write_csv(fig_2_clean_data, "project-files/data-clean/fig_2_clean_data.csv")
```


# Visualization 1

> Here is a description of the first visualization. I want to describe what the visualization shows to a general audience. I do not want to see any code or technical terms. I want to be able to read this and understand what the visualization shows. I also want to see a caption and alt text for the visualization. The caption should be a short description of the visualization and the alt text should be a longer description of the visualization. The alt text should be descriptive enough that someone who is blind could understand what the visualization shows.

```{r}
#| fig-cap: "Fig. 1: Yearly mean scores for each of the 6 International Math Olympiad questions. Each question is scored from 1 to 7 points. "
#| fig-alt: "This is an alternative text for the first visualization"
#| fig-width: 6
#| fig-height: 4
#| fig-align: center
fig_1_clean_data %>% 
  ggplot(aes(x = year, y = mean_score)) + 
  geom_line() +
  facet_grid(rows = vars(problem_num)) +
  labs(x = "Year",
       y = "Mean Score") +
  theme_light()
```

# Visualization 2

> Here is a description of the second visualization. I want to describe what the visualization shows to a general audience. I do not want to see any code or technical terms. I want to be able to read this and understand what the visualization shows. I also want to see a caption and alt text for the visualization. The caption should be a short description of the visualization and the alt text should be a longer description of the visualization. The alt text should be descriptive enough that someone who is blind could understand what the visualization shows.

```{r}
#| fig-cap: "Fig. 2: Yearly mean scores per continent for each of the 6 International Math Olympiad questions. Each question is scored from 1 to 7 points. "
#| fig-alt: "This is an alternative text for the second visualization"
#| fig-width: 6
#| fig-height: 4
#| fig-align: center
fig_2_clean_data %>% 
  ggplot(aes(x = year, y = mean_score)) + 
  geom_line() +
  facet_grid(rows = vars(problem_num), cols = vars(continent)) +
  labs(x = "Year",
       y = "Mean Score") +
  theme_light() +
  theme(axis.text.x = element_text(angle = 60, hjust = 1))
  
```
